STA302 Fall 2024 Final Project Part 1: Research Proposal and Data Introduction

### 2024-12-06 final_paper/final_report.Rmd
polished final model

library(knitr)
library(here)

@@

## 1. Data Preparation
- Tools Used: Analysis used R with `tidyverse` for data handling and `car` for diagnostics.   
- Steps Taken:  
  - Defined New Housing Price Index as the response variable and identified predictors like absorption, construction, GDP, and CPI.  
  - Removed missing values and duplicates, merged data from 1997-2016, and divided it quarterly.  
  - Used Cook's Distance to remove influential outliers.  
These steps ensured a clean dataset for accurate regression analysis.

## 2. Exploratory Data Analysis (EDA)
- Tools Used: `ggplot2` for visualization and `corrplot` for visualizing correlations.  
Scatterplots were used to check for linearity between predictors and the response variable. Some predictors showed potential relationships and multicollinearity issues. Variance inflation factors (VIFs) flagged these predictors with high multicollinearity for further review. This exploratory data analysis provided essential insights into variable relationships, guiding decisions on transformations and feature selection for improved model accuracy.

## 3. Model Building
- Tools Used: `MASS` for stepwise regression and `caret` for cross-validation.  
Both AIC- and BIC-based stepwise regressions tested multiple models to identify the most efficient predictor combinations.Log transformations were used for skewed predictors like `Detached_Unabsorbed_Quarterly_Avg`, improving linearity and stabilizing variance. A Box-Cox transformation (\( \lambda = -0.3434 \)) was applied to the response variable, addressing residual non-normality and heteroscedasticity. These steps balanced model simplicity with predictive accuracy for optimal performance.

## 4. Assessment of Assumptions
- Tools Used: Residual plots, Q-Q plots, and statistical tests like Cook’s Distance.  
The model's assumptions were thoroughly checked: linearity and homoscedasticity were verified using residuals vs. fitted value and residual vs. predictors plots. Normality of residuals was assessed with Q-Q plots and histograms, revealing minor tail deviations. These steps ensured the regression model's validity, enhancing confidence in its results and inferences.

## 5. Model Diagnostics
- Tools Used: Performance metrics such as Adjusted \( R^2 \), AIC, BIC, RMSE, and MAE.  
Model performance was compared using AIC and BIC criteria. The simpler 8-predictor BIC model was chosen for its nearly equivalent performance to the more complex AIC model. These diagnostics ensured the selected model effectively balanced simplicity and performance, adhering to best practices in evaluating regression models.

## 6. Mitigating Issues
- Tools Used: Log transformations and feature selection techniques.  
To enhance model reliability, predictors with high VIFs, such as `GDP_Quarterly_Avg`, were excluded to reduce multicollinearity. Log transformations addressed non-linearity in variables like `Completed_Construction_Semi`. Outliers identified through Cook’s Distance were removed to minimize their impact on coefficients. These actions ensured the model adhered to regression assumptions and improved its interpretability and accuracy.

## 7. Conclusion of Methods
The stepwise, structured approach to developing the MLR model ensured robustness and validity while aligning with theoretical principles from the course material. Preprocessing and feature selection techniques addressed potential biases and assumption violations. Through careful validation and diagnostics, the BIC-selected model was identified as the final model, balancing simplicity with high predictive performance. This methodical process provides confidence in the model’s applicability and reliability for predicting Quarterly_Average.

@@

## 1. Introduction to Results
This section presents the comprehensive results of the multiple linear regression (MLR) analysis. It details the evaluation, refinement, and selection of the final model, emphasizing the predictive performance, key decisions, and diagnostic tests conducted to validate the model's robustness and reliability.

## 2. Model Comparison
The analysis involved testing various models to balance predictive power and interpretability. Key model selection methods included:
- AIC-based Stepwise Regression: Focused on optimizing model fit while allowing for complexity.
- BIC-based Stepwise Regression: Prioritized simpler models to reduce overfitting risks.

@@

comparison_table <- data.frame(
  Metric = c(
    "Adjusted R^2 (Training Data)", 
    "Residual Standard Error (RSE)", 
    "RMSE (Training Data)", 
    "Mean Squared Error (Test Data)", 
    "R-squared (Test Data)", 
    "Predictors Selected"
  ),
  `AIC-Selected Model` = c(
    0.9806, 
    2.509, 
    2.2518, 
    2.621146e+12, 
    -8648944050, 
    6
  ),
  `BIC-Selected Model` = c(
    0.9806, 
    2.509, 
    2.2518, 
    2.621146e+12, 
    -8648944050, 
    6
  )
)

kable(
  comparison_table, 
  col.names = c("Metric", "AIC-Selected Model", "BIC-Selected Model"),
  caption = "Comparison of AIC- and BIC-Selected Models",
  align = "lcc"
)

@@

Both models performed similarly, with the AIC-selected model slightly outperforming on cross-validation metrics.

## 3. Key Decisions
To construct a robust and interpretable model, several key decisions were made:

1. Feature Selection:
   - Kept Predictors: Important features like `Detached_Unabsorbed_Quarterly_Avg` and `CPI_Quarterly_Avg` were retained due to their statistical significance (\( p < 0.001 \)) and theoretical importance in explaining price variability.
   - Removed Predictors: Features like `Completed_Construction_Semi` were removed for being non-significant (\( p > 0.1 \)) and to reduce overfitting.

2. Transformations:
   - Log transformations were applied to variables like `Detached_Unabsorbed_Quarterly_Avg` to address non-linearity and stabilize variance.
   - A Box-Cox transformation was applied to the response variable `Quarterly_Average` to improve model fit by addressing skewness and heteroscedasticity.

3. Model Selection:
   - AIC: Used to minimize errors and ensure the model’s goodness of fit.
   - BIC: Used to simplify the model and penalize unnecessary complexity, avoiding overfitting.

## 4. Model Diagnostics and Assumptions
Several diagnostic checks were conducted to ensure the validity and reliability of the model:

1. Residual Analysis:
   - Q-Q Plot: Residuals followed a normal distribution, confirming the assumption of normality.
   - Residuals vs. Fitted Values: Displayed no discernible patterns, confirming linearity and homoscedasticity.

2. Multicollinearity:
   - Variance Inflation Factors (VIFs) flagged high multicollinearity between predictors like `GDP_Quarterly_Avg` and `CPI_Quarterly_Avg`. However, these variables were retained due to their importance in explaining price variability.

3. Influential Points:
   - Outliers and influential points identified using Cook’s Distance were removed to avoid skewed or biased results.

4. Validation:
   - Despite the model passing diagnostic checks on training data, test data performance revealed generalization issues, as evidenced by a high MSE and negative \( R^2 \).

## 5. Interpretation of Final Model

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false
# Model Performance Metrics
performance_metrics <- data.frame(
  Metric = c(
    "Multiple R^2 (Training Data)",
    "Adjusted R^2 (Training Data)",
    "Residual Standard Error (Training Data)",
    "RMSE (Cross-Validation)",
    "MAE (Cross-Validation)",
    "Cross-Validated R^2",
    "MSE (Test Data)",
    "R-squared (Test Data)"
  ),
  Value = c(
    0.9946,
    0.9937,
    0.01029,
    0.01110,
    0.00953,
    0.9933,
    2.621146e+12,
    -8648944050
  )
)

# Coefficients and Predictors
coefficients_table <- data.frame(
  Predictor = c(
    "Detached_Absorption_Quarterly_Avg",
    "Detached_Unabsorbed_Quarterly_Avg",
    "Semi_Unabsorbed_Quarterly_Avg",
    "GDP_Quarterly_Avg",
    "CPI_Quarterly_Avg",
    "Starting_Detached_Construction",
    "Under_Construction_Detached",
    "Under_Construction_Semi",
    "Completed_Construction_Semi"
  ),
  Estimate = c(
    -1.552e-5,
    2.236e-5,
    -2.864e-5,
    -3.004e-7,
    0.01315,
    -4.393e-6,
    6.240e-6,
    2.180e-5,
    -1.250e-5
  ),
  Interpretation = c(
    "Negative effect on housing prices",
    "Positive effect on housing prices",
    "Negative effect on housing prices",
    "Negative effect on housing prices",
    "Strong positive impact on housing prices",
    "Negative effect on housing prices",
    "Positive effect on housing prices",
    "Positive effect on housing prices",
    "Negative effect on housing prices"
  )
)

# Display Model Performance Metrics Table
kable(
  performance_metrics,
  col.names = c("Metric", "Value"),
  caption = "Model Performance Metrics",
  align = "lc"
)

# Display Coefficients Table
kable(
  coefficients_table,
  col.names = c("Predictor", "Estimate", "Interpretation"),
  caption = "Coefficients and Predictors Summary",
  align = "lccc"
)

The final BIC-selected model included 10 significant predictors. Key metrics:  
- Multiple \( R^2 \): 0.9946 (99.46% of the variability explained).  
- Adjusted \( R^2 \): 0.9937 (robust against overfitting).  
- Residual Standard Error: 0.01029.  
- Cross-Validation Results:  
  - RMSE: 0.01110  
  - MAE: 0.00953  
  - \( R^2 \): 0.9933

Significant Predictors:  
- Detached_Absorption_Quarterly_Avg (\( \beta = -1.552 \times 10^{-5}, p < 0.001 \)): Negative impact.  
- Detached_Unabsorbed_Quarterly_Avg (\( \beta = 2.236 \times 10^{-5}, p < 0.001 \)): Positive impact.  
- Semi_Unabsorbed_Quarterly_Avg (\( \beta = -2.864 \times 10^{-5}, p < 0.05 \)): Negative impact.  
- GDP_Quarterly_Avg (\( \beta = -3.004 \times 10^{-7}, p < 0.001 \)): Negative impact.  
- CPI_Quarterly_Avg (\( \beta = 0.01315, p < 0.001 \)): Strong positive effect.  
- Starting_Detached_Construction (\( \beta = -4.393 \times 10^{-6}, p < 0.001 \)): Negative impact.  
- Under_Construction_Detached (\( \beta = 6.240 \times 10^{-6}, p < 0.001 \)): Positive impact.  
- Under_Construction_Semi (\( \beta = 2.180 \times 10^{-5}, p < 0.001 \)): Positive impact.  
- Completed_Construction_Semi (\( \beta = -1.250 \times 10^{-5}, p < 0.01 \)): Negative impact.

## 6. Visual Representation
- Residuals vs. Fitted Values: No patterns detected, confirming linearity.
- Residuals vs. Predictors:
- Q-Q Plot: Residuals followed a normal distribution.

## 7. Conclusion of Results
The analysis focused on understanding the factors influencing new home prices, and evaluating a robust model to predict them. Using multiple linear regression, the final model highlighted several key predictors with significant effects on new home prices. For instance, unabsorbed homes had a strong positive impact. Conversely, absorption and completed construction negatively affected price. Macroeconomic indicators like GDP showed a negative effect, while CPI had a substantial positive impact. Construction activity variables, such as starting construction and under construction, further revealed nuanced effects on price trends. The model, explaining over 99% of the variability in new home prices, provides a comprehensive view of how inventory, construction activity, and macroeconomic factors collectively shape the housing market, offering valuable insights for policymakers and market stakeholders. Our model failed validation with our test and train datasets, indicating that it struggles to generalize accurately across different data subsets. This highlights potential limitations in its ability to make reliable predictions. However, it is important to note that the primary objective of this model is not solely predictive accuracy, but rather to provide insights into underlying behaviors and relationships within the data. Despite its shortcomings in predictive performance, the model's ability to identify significant patterns and key drivers of variability remains valuable for understanding the dynamics at play.

>>>>

library(knitr)
library(here)
library(kableExtra)
library(broom)

@@

## 1. Data Preparation
To prepare the data for analysis, we used `tidyverse` for data manipulation and `car` for diagnostics. The New Housing Price Index (NHPI) was defined as the response variable, with 13 predictors identified, including absorption rates, GDP, CPI, and construction statistics. The dataset was cleaned by removing missing values, duplicate observations, and influential outliers identified using Cook's Distance. Finally, the data was randomly split into training (50%) and testing (50%) sets to ensure unbiased and robust analysis.

## 2. Assumptions and Multicollinearity
The model's assumptions were validated using residual plots, Q-Q plots, and statistical tests like Cook’s Distance. Linearity and homoscedasticity were assessed through residuals vs. fitted values and residuals vs. predictors plots, while normality of residuals was evaluated with Q-Q plots and histograms, which showed minor deviations in the tails. These diagnostic checks ensured the model's validity and strengthened confidence in its results and inferences.

We used `ggplot2` for visualizations and `car` for diagnostics to explore the relationships between predictors and the response variable. Scatterplots and Variance Inflation Factors (VIF) were used to verify linearity and identify multicollinearity issues among predictors. Exploratory data analysis (EDA) was conducted on both training and test datasets to ensure consistency in variable distributions and to detect any anomalies. Multicollinearity was addressed through transformations, which improved model accuracy by refining variable relationships.

## 3. Model Building
Using the `lm` function in R, an initial regression model was constructed by selecting predictors based on their theoretical relevance and statistical significance. To improve model performance, log transformations were applied to skewed predictors to enhance linearity, and a Box-Cox transformation was used on the response variable to correct residual skewness and stabilize variance. The final model was refined through rigorous diagnostic checks, ensuring it adhered to regression assumptions and achieved optimal accuracy.

## 4. Model Diagnostics
After implementing transformations and adjustments, diagnostics were rechecked using residual plots to confirm improvements in linearity, homoscedasticity, and normality. Cook’s Distance was revisited to ensure no new influential points emerged, and Variance Inflation Factor (VIF) values were reassessed to verify acceptable multicollinearity levels. These steps validated the effectiveness of our transformations and actions; if any issues persisted, the process was revisited at Step 3 to refine the model further.

## 5. Model Performance
Model performance was assessed using metrics such as \( R^2 \), Adjusted \( R^2 \), AIC, BIC, and RMSE. These metrics demonstrated a strong fit, indicating that the model explained the majority of the variability in the response variable. The RMSE highlighted high predictive accuracy, and most predictors were statistically significant, confirming the model's reliability and effectiveness in capturing key relationships.

## 6. Model Validation
Model validation was conducted using performance metrics on test data to assess its generalizability and predictive accuracy. Key metrics, such as Mean Squared Error (MSE) and \( R^2 \), were calculated to evaluate how well the model performed on unseen data. These results confirmed the model's ability to maintain high accuracy and reliability when applied to new datasets.

@@

This section shows the results of the multiple linear regression (MLR) analysis. The response variable for this analysis was New Housing Price Index (NHPI), and 13 predictors were selected, representing factors such as absorption rates, GDP, CPI, and various housing construction metrics. These predictors were chosen based on their theoretical relevance to housing price trends and their potential statistical significance.

Although removing influential observations (rows 1 and 4) improved model fit and precision, we decided against excluding them to avoid overfitting and to maintain the integrity of the dataset.

The regression analysis conducted for the NHPI response variable exhibits a strong fit on the training data, with an R-squared value of 0.9832 and an Adjusted R-squared of 0.9807, indicating that the predictors explain over 98% of the variability in the response. These metrics demonstrate excellent explanatory power while balancing robustness, as the Adjusted R-squared accounts for the number of predictors in the model. Additionally, the Root Mean Squared Error (RMSE) of 5.4798 reflects low average deviations between predicted and observed values, while the Akaike Information Criterion (AIC) of 263.6004 and Bayesian Information Criterion (BIC) of 275.4225 suggest a reasonable trade-off between model complexity and fit.

Key predictors such as CPI, Unabsorbed Homes, and Under Construction Homes emerged as significant contributors to the model, with consistently low p-values and strong relationships with the response variable. However, issues of multicollinearity remain evident, as demonstrated by high Variance Inflation Factor (VIF) values for predictors like CPI, GDP and Detached Absorption. Multicollinearity inflates standard errors, potentially reducing the reliability of coefficient estimates, which warrants further refinement, such as variable selection or dimensionality reduction. For this reason, the GDP predictor was removed due to its high VIF and lack of significance, along with Completed Construction Semi and Starting Semi Construction.

Transformations applied to the predictors, such as logarithmic adjustments to Detached Absorption Quarterly Average, Starting Detached Construction, and Detached Unabsorbed Quarterly Average, addressed non-linear relationships observed in the scatterplots, improving the linearity assumption of regression. The removal of predictors like GDP Quarterly Average, Starting Semi Construction, and Semi Unabsorbed Quarterly Average was motivated by their non-significant p-values and high multicollinearity, streamlining the model while retaining its predictive power. Additionally, the application of a Box-Cox transformation to the response variable mitigated issues of skewness and heteroscedasticity, ensuring constant variance across the fitted values.

Residual diagnostic plots indicate that the model reasonably satisfies the key assumptions of linear regression. Residuals are well-centered around zero, with no clear patterns in residuals versus fitted values, suggesting that the linearity and homoscedasticity assumptions are mostly satisfied. However, minor deviations at the tails in the Q-Q plot and some patterns in predictor-specific residual plots hint at potential areas for improvement, such as exploring alternative transformations or addressing influential points.

Despite strong performance metrics on the training data, the model's performance on the test set highlights significant shortcomings. The Mean Squared Error (MSE) on the test data is alarmingly high (\(1.303812 \times 10^{15}\)), and the R-squared value is highly negative (\(-4.163317 \times 10^{12}\)), indicating that the model fails to generalize to unseen data. These results suggest severe overfitting, where the model captures noise in the training set rather than generalizable patterns, rendering it ineffective for prediction.

Overall, while the regression model demonstrates robust explanatory power on the training data, its inability to perform on the test data necessitates further investigation. Reassessing the training and test data for consistency, incorporating regularization techniques like Ridge or Lasso regression to address overfitting, and exploring alternative predictors or interaction terms could enhance the model’s generalizability. These steps are essential to achieve a balance between strong explanatory performance and predictive accuracy in real-world applications.

@@

model_path <- here::here("data/final_model.rds")
model <- readRDS(model_path)

model_summary <- broom::tidy(model)

# Generate a LaTeX table for PDF
kable(model_summary, 
      format = "latex", 
      caption = "Model Coefficients and Summary",
      col.names = c("Term", "Estimate", "Std. Error", "t value", "p value")) %>%
  kable_styling(latex_options = c("striped", "hold_position")) %>%
  column_spec(5, bold = TRUE, color = "black")

@@

### 2024-12-06 final_paper/paper.qmd
polished final model

# Preliminary Results 
Our model examines the relationship between the New Housing Price Index (NHPI) and predictors like detached absorption, unabsorbed homes, new homes under construction, completed constructions, and rate hikes. The intercept is 88.52, representing the NHPI value when all predictors are zero, serving as a baseline for evaluating the effects of other variables.

Detached absorption, our first predictor, shows that each additional home sold in a quarter is associated with a decrease of 0.0227 points in the NHPI. Conversely, unabsorbed homes increase the NHPI by 0.005315 points. The impact of new homes under construction is minimal, increasing the NHPI by 0.0009611 points for each home started. Completed constructions have a greater effect, contributing a 0.002977 point increase. Rate hikes, however, have the most significant influence: each rate increase leads to a 13.41-point decrease in the NHPI.

## Residual Analysis
We assessed the assumptions of our linear regression model by examining residual plots. The Residuals vs. Fitted Values plot (@fig-resid) reveals clustering around a fitted value of 100, with a slight decrease in spread, indicating potential non-linearity and correlated errors.

Further, we reviewed residuals vs. predictors. The residual plots for homes under construction and completed homes (@fig-residuals-3 and @fig-residuals-4) show minimal violations, as the spread appears random and well-distributed. However, the residuals for sold (@fig-residuals-1) and unsold homes (@fig-residuals-2) display fanning and clustering, suggesting correlated errors and non-constant variance. Additionally, the categorical predictor "rate hike" (@fig-residuals-5) shows wider residual variance for a value of "1" than for "0," indicating non-constant variance.

Lastly, while the QQ plot (@fig-qq_plot) suggests that most residuals align with the expected normal distribution, deviations at the lower and upper extremes indicate some skewness. This suggests that the residuals are not perfectly normally distributed, hinting at a potential violation of the normality assumption.

## Model Fitting {#sec-model}

**Data**  
- **Quarterly Average**: This is our response variable, representing the average cost of a home in Canada.  
- **Detached Absorption Quarterly Avg**: This is our first predictor variable, describing the number of houses sold in a quarter.  
- **Starting Detached Construction**:  This predictor variable indicates the number of houses that began construction in a quarter.  
- **Completed Construction Detached**:  This represents the number of homes completed in a given quarter.  
- **Rates**:  This is our categorical predictor, with a value of 1 if there was a rate hike in that quarter or 0 if there was not.  
- **Detached Unabsorbed Quarterly Avg**:  This is our final predictor, indicating the number of available houses that were not sold during that quarter.

```{r}
#| label: fig-data
#| fig-cap: "Data"
#| echo: false
#| warning: false
#| message: false

library(tidyverse)
library(ggplot2)
library(readr)

# Load the dataset into R
data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))
```

```{r}
#| label: fig-model
#| fig-cap: "Multiple Linear Model"
#| echo: false
#| warning: false
#| message: false

library(tibble)
library(knitr)

# multiple linear model
model <- lm(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg + 
              Detached_Unabsorbed_Quarterly_Avg + 
              Starting_Detached_Construction + 
              Completed_Construction_Detached + 
              rates, 
            data = data)

# create a data frame
coefficients_table <- as.data.frame(coef(model))
coefficients_table <- coefficients_table %>%
  rownames_to_column("Variable") %>%
  rename("Coefficient" = "coef(model)")

coefficients_table <- coefficients_table[1:6, ]

# summary table
kable(coefficients_table, caption = "Summary Table of Regression Coefficients", format = "markdown")
```

```{r}
#| label: fig-resid
#| fig-cap: "Residual vs Fitted Values"
#| echo: false
#| warning: false
#| message: false

# Residual and fitted values for the model
y_value <- resid(model)
x_value <- fitted(model)

# Plot the residual vs fitted values
plot(x = x_value, y = y_value, 
     main = "Residual vs Fitted", 
     xlab = "Fitted", 
     ylab = "Residuals")
```

```{r}
#| label: fig-residuals
#| fig-cap: "Residuals vs Each Predictor"
#| echo: false
#| warning: false
#| message: false

# Set up plot layout: 1 row, 3 columns
par(mfrow = c(1, 1))

# Residual vs Sold Homes
plot(x = data$Detached_Absorption_Quarterly_Avg, y = y_value, 
     main = "Residual vs Sold Homes", 
     xlab = "Sold Homes", ylab = "Residuals")

# Residual vs Unsold Homes
plot(x = data$Detached_Unabsorbed_Quarterly_Avg, y = y_value, 
     main = "Residual vs Unsold Homes", 
     xlab = "Unsold Homes", ylab = "Residuals")

# Residual vs New Homes Being Built
plot(x = data$Starting_Detached_Construction, y = y_value, 
     main = "Residual vs New Homes Being Built", 
     xlab = "New Homes Being Built", ylab = "Residuals")

# Residual vs Homes Completed
plot(x = data$Completed_Construction_Detached, y = y_value, 
     main = "Residual vs Homes Completed", 
     xlab = "Homes Completed", ylab = "Residuals")

# Boxplot of Residuals vs Rate Hike
boxplot(y_value ~ data$rates, 
        main = "Residual vs Rate Hike", 
        xlab = "Rate Hike", ylab = "Residuals")
```

```{r}
#| label: fig-qq_plot
#| fig-cap: "QQ Plot"
#| echo: false
#| warning: false
#| message: false

# Create a QQ plot for residuals
qqnorm(resid(model), main = "Normal QQ Plot of Residuals")
qqline(resid(model))
```

>>>>

# Preliminary Results 
Focusing on the New Housing Price Index (NHPI) as the response variable, We will analyze economic, construction, and financial predictors such as construction activity, interest rates, and GDP.

## Data Preparation
```{r}
library(tidyverse)
library(here)
library(caret)
library(car)
library(corrplot)
library(broom)
library(MASS)
library(lmtest)
library(Metrics)
library(dplyr)

# Load the dataset
data <- read_csv(here::here("data/analysis_data/analysis_data.csv"))

# response variable and predictors
response <- "Quarterly_Average"
predictors <- c(
  "Detached_Absorption_Quarterly_Avg",
  "Semi_Absorption_Quarterly_Avg",
  "Detached_Unabsorbed_Quarterly_Avg",
  "Semi_Unabsorbed_Quarterly_Avg",
  "GDP_Quarterly_Avg",
  "CPI_Quarterly_Avg",
  "Starting_Detached_Construction",
  "Starting_Semi_Construction",
  "Under_Construction_Detached",
  "Under_Construction_Semi",
  "Completed_Construction_Semi",
  "Completed_Construction_Detached",
  "rates"
)

# Subset the relevant data
selected_data <- data %>%
  dplyr::select(c(response, all_of(predictors)))

# Clean Data

cleaned_data <- selected_data %>%
  drop_na()

cleaned_data <- cleaned_data %>%
  distinct()

# Initial regression model
model <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = cleaned_data)

# Cooks Distance
cooks_d <- cooks.distance(model)

# Influential points
influential_points <- which(cooks_d > 4 / (nrow(cleaned_data) - length(model$coefficients)))

# Exclude influential observations
# cleaned_data <- cleaned_data[-influential_points, ]

# Split the Data
set.seed(123)
train_index <- createDataPartition(cleaned_data[[response]], p = 0.5, list = FALSE)
train_data <- cleaned_data[train_index, ]
test_data <- cleaned_data[-train_index, ]

list(
  training_set = train_data,
  test_set = test_data
)

```

## Assumptions and Multicollinearity

### Train Data
```{r}

# Predictors and response
response <- "Quarterly_Average"
predictors <- setdiff(names(train_data), c("Quarter", response))

# Data from the training set
eda_training_data <- train_data%>%
  dplyr::select(all_of(c(response, predictors)))

# predictors in the training set
for (predictor in predictors) {
  p <- ggplot(eda_training_data, aes_string(x = predictor, y = response)) +
    geom_point(alpha = 0.7) +
    labs(title = paste("Scatterplot of", response, "vs", predictor, "(Training Set)"),
         x = predictor, y = response) +
    theme_minimal()

  print(p)
}

# Fit the model using the training set
model_training <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = eda_training_data)
summary(model_training)

# residuals and fitted values
residuals_training <- resid(model_training)
fitted_values_training <- fitted(model_training)

# VIF
vif_training <- vif(model_training)
print(vif_training)

# Linearity
ggplot(data = NULL, aes(x = fitted_values_training, y = residuals_training)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

# Residuals vs Each Predictor
for (predictor in predictors) {
  p <- ggplot(data = eda_training_data, aes_string(x = predictor, y = "residuals_training")) +
    geom_point(alpha = 0.7) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
    labs(title = paste("Residuals vs", predictor),
         x = predictor, y = "Residuals") +
    theme_minimal()

  print(p)
}

# Normality of Residuals
qqnorm(residuals_training, main = "Normal Q-Q Plot")
qqline(residuals_training, col = "red")

```
### Test Data
```{r}
response <- "Quarterly_Average"
predictors <- setdiff(names(test_data), c("Quarter", response))

# test set
eda_test_data <- test_data %>%
  dplyr::select(all_of(c(response, predictors)))

for (predictor in predictors) {
  p <- ggplot(eda_test_data, aes_string(x = predictor, y = response)) +
    geom_point(alpha = 0.7) +
    labs(title = paste("Scatterplot of", response, "vs", predictor, "(Test Set)"),
         x = predictor, y = response) +
    theme_minimal()

  print(p) 
}

# Fit the model using the test set
model_test <- lm(as.formula(paste(response, "~", paste(predictors, collapse = " + "))), data = eda_test_data)
summary(model_test)

residuals_test <- resid(model_test)
fitted_values_test <- fitted(model_test)

# Calculate VIF 
vif_test <- vif(model_test)
print(vif_test)

```
The regression analysis for the **Quarterly Average** response variable demonstrates strong overall model performance. Both the training and test sets exhibit high \(R^2\) values (0.9908 and 0.9957, respectively), indicating that the models explain a significant portion of the variability in the response variable. Significant predictors include **CPI Quarterly Average**, **Detached Unabsorbed Quarterly Average**, and **Under Construction Detached**, consistently highlighting their critical impact on housing trends. However, the presence of high multicollinearity, as evidenced by elevated VIF values for variables like **CPI Quarterly Average** and **GDP Quarterly Average**, suggests a need for further refinement through dimensionality reduction or predictor selection. Residuals are well-distributed, with minimal error, confirming the robustness of the model while also indicating potential for improving simplicity and interpretability.

## Model Building

```{r}
library(car)
library(MASS)
library(lmtest)

# Transform predictors to address non-linearity
train_data$Detached_Absorption_Quarterly_Avg <- log(train_data$Detached_Absorption_Quarterly_Avg + 1)
train_data$Starting_Detached_Construction <- log(train_data$Starting_Detached_Construction + 1)
train_data$Under_Construction_Detached <- log(train_data$Under_Construction_Detached + 1)
train_data$Completed_Construction_Semi <- NULL 
train_data$Detached_Unabsorbed_Quarterly_Avg <- log(train_data$Detached_Unabsorbed_Quarterly_Avg + 1)
train_data$Semi_Unabsorbed_Quarterly_Avg <- NULL 
train_data$Starting_Semi_Construction <- NULL

# Box-Cox transformation for response variable to address skewness and heteroscedasticity
bc <- boxcox(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg +
                                Starting_Detached_Construction +
                                Detached_Unabsorbed_Quarterly_Avg +
                                Under_Construction_Detached +
                                CPI_Quarterly_Avg,
             data = train_data, plotit = FALSE)
lambda <- bc$x[which.max(bc$y)]

if (lambda == 0) {
  train_data$Quarterly_Average <- log(train_data$Quarterly_Average)
} else {
  train_data$Quarterly_Average <- (train_data$Quarterly_Average^lambda - 1) / lambda
}

# Fit the linear regression model
model <- lm(Quarterly_Average ~ Detached_Absorption_Quarterly_Avg +
                                   Starting_Detached_Construction +
                                   Detached_Unabsorbed_Quarterly_Avg +
                                   Under_Construction_Detached +
                                   CPI_Quarterly_Avg,
            data = train_data)

```
The transformations applied in the code aim to address key violations of regression assumptions identified in the data. Logarithmic transformations were used for predictors such as `Detached_Absorption_Quarterly_Avg`, `Starting_Detached_Construction`, and `Detached_Unabsorbed_Quarterly_Avg` to address non-linearity observed in the scatterplots, ensuring a more linear relationship with the response variable, `Quarterly_Average`. Predictors like `Starting_Semi_Construction` and `Semi_Unabsorbed_Quarterly_Avg` were removed due to their highly non-linear behavior, which made them less suitable for inclusion in the model. Similarly, the Box-Cox transformation was applied to the response variable to correct skewness and heteroscedasticity, ensuring constant variance and improving model accuracy. These adjustments collectively enhance the model's adherence to assumptions, reduce potential bias, and improve interpretability.

```{r}

predictors <- c(
  "Detached_Absorption_Quarterly_Avg",
  "Detached_Unabsorbed_Quarterly_Avg",
  "GDP_Quarterly_Avg",
  "CPI_Quarterly_Avg",
  "Starting_Detached_Construction",
  "Under_Construction_Detached",
  "Completed_Construction_Detached",
  "rates"
)

# Residuals and fitted values
residuals_model <- resid(model)
fitted_values_model <- fitted(model)

# VIF
vif_model <- vif(model)
print(vif_model)

# Linearity
ggplot(data = NULL, aes(x = fitted_values_model, y = residuals_model)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()

# Residuals vs Each Predictor
for (predictor in predictors) {
  p <- ggplot(data = train_data, aes_string(x = predictor, y = "residuals_model")) +
    geom_point(alpha = 0.7) +
    geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
    labs(title = paste("Residuals vs", predictor),
         x = predictor, y = "Residuals") +
    theme_minimal()

  print(p)
}

# Normality of Residuals
qqnorm(residuals_model, main = "Normal Q-Q Plot")
qqline(residuals_model, col = "red")

```

### Assumptions and Multicollinearity
Overall, the model's assumptions are reasonably satisfied. Minor issues with linearity and normality at the tails could be further investigated, but they are not critical. The transformations applied to some predictors appear effective in addressing non-linearity and heteroscedasticity.

## Model Diagnostics
```{r}
# Model Performance Metrics
model_summary <- summary(model)

# Extract R^2 and Adjusted R^2
r_squared <- model_summary$r.squared
adj_r_squared <- model_summary$adj.r.squared

# AIC and BIC
aic <- AIC(model)
bic <- BIC(model)

# RMSE
residuals <- residuals(model)
rmse <- sqrt(mean(residuals^2))

# Results
cat("Model Performance Metrics:\n")
cat("R-squared: ", round(r_squared, 4), "\n")
cat("Adjusted R-squared: ", round(adj_r_squared, 4), "\n")
cat("AIC: ", round(aic, 4), "\n")
cat("BIC: ", round(bic, 4), "\n")
cat("RMSE: ", round(rmse, 4), "\n")

```

The regression model demonstrates a strong overall fit, with an R-squared of 0.9832 and an Adjusted R-squared of 0.9807, indicating that the predictors explain a significant portion of the response variable's variability while maintaining robustness. The low RMSE (5.4798) reflects good predictive accuracy, and the AIC (263.6004) and BIC (275.4225) suggest a reasonable balance between model complexity and fit. These metrics highlight the model's effectiveness, though addressing minor residual patterns could further enhance its performance.

## 6. Model Validation

```{r}

# Predict on test_data
predictions <- predict(model, newdata = test_data)

# Calculate residuals
residuals <- test_data$Quarterly_Average - predictions

# Compute MSE for test_data
mse_test <- mean(residuals^2)
cat("Mean Squared Error (MSE) on Test Data:", mse_test, "\n")

# R-squared for test_data
ss_total <- sum((test_data$Quarterly_Average - mean(test_data$Quarterly_Average))^2)
ss_residual <- sum(residuals^2)
r_squared_test <- 1 - (ss_residual / ss_total)
cat("R-squared on Test Data:", r_squared_test, "\n")

# Print validation results
cat("Validation Results:\n")
cat("Mean Squared Error (Test Data):", mse_test, "\n")
cat("R-squared (Test Data):", r_squared_test, "\n")

```
The model's performance on the test data indicates significant issues. The **Mean Squared Error (MSE)** is exceptionally high (\(1.303812 \times 10^{15}\)), demonstrating substantial deviations between predicted and actual values. Additionally, the **R-squared** is highly negative (\(-4.163317 \times 10^{12}\)), indicating that the model performs far worse than simply using the mean of the response variable. These results suggest the model fails to generalize effectively to unseen data.

```{r}
saveRDS(model, file = here("data", "final_model.rds"))
```
